Higher-Order unification for free!

### 2024-04-16 paper/main.tex
zap ML from the abstract

\begin{abstract}
  Specifying and implementing a proof system from scratch requires significant effort.
  Logical Frameworks and Higher Order Logic Programming Languages provide
  dedicated, high-level Meta Languages (ML) to facilitate this task in two
  key ways: 1) variable binding and substitution are for free when ML binders
  represent object logic ones; 2) proof construction, and even proof search, are
  greatly simplified by leveraging the unification procedure provided by the ML.
  Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
  $\lambda$Prolog~\cite{miller_nadathur_2012} and
  Isabelle~\cite{10.1007/978-3-540-71067-7_7}
  which have been utilized to implement various formal systems such as
  First Order Logic~\cite{felty88cade},
  Set Theory~\cite{10.1007/BF00881873},
  Higher Order Logic~\cite{books/sp/NipkowPW02}, and even the Calculus of
  Constructions~\cite{felty93lics}.

  The object logic we are interested in is
  Coq's Calculus of Inductive Constructions (CIC)\cite{Coq-refman}. We aim to
  develop a higher-order unification-based 
  proof search procedure for it using the ML
  Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
  Elpi's equational theory includes
  $\eta\beta$ equivalence and features a
  higher-order unification procedure \Ue restricted to the pattern
  fragment~\cite{miller92jsc}.
  Elpi offers an encoding of CIC suitable
  for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}
  but restricts \Ue to roughly first-order unification problems only.
  We refer to this basic encoding as \Fo.

  In this paper we propose a more well-behaved encoding called \Ho,
  and show how to translate unification problems from \Fo{}
  to corresponding ones in \Ho.
  Consequently, we derive \Uo{}, the higher-order unification procedure of \Fo{}
    that honours $\eta\beta$-equivalence (for CIC functions), addresses
  problems within the pattern fragment, and allows
  for the use of heuristics to deal with problems outside the
  pattern fragment. Moreover, as \Uo{} delegates most of the work to \Ue,
  it can be used to efficiently simulate a logic program in \Fo{} by
  taking advantage of
  unification-related optimizations of the ML, such as clause indexing.

@@

\begin{proposition}[Compilation round trip]\label{prop:comprt}
  % \elpiIn{compile F T [] M [] L [] S} then 
  % if \elpiIn{T} $\in$ S
  If $\C{s}{t}{m}{l}$, such that $l \in \linkStore$ and $m \in \mapStore$, if
  $\sigma = \{ A \mapsto t\}$ and $\mapping{X}{A}{n} \in \mapStore$
  then $\D{\sigma}{\mapStore}{\linkStore}{\rho}$ and
  $\rho X \Eo \rho s$.
\end{proposition}
\todo{explain no proof, llam link breaks}

>>>>

\begin{abstract}
  Specifying and implementing a proof system from scratch requires significant effort.
  Logical Frameworks and Higher Order Logic Programming Languages provide
  dedicated, high-level meta languages to facilitate this task in two
  % key 
  ways: 1) variable binding and substitution are for free when meta language binders
  represent object logic ones; 2) proof construction, and %even 
  proof search, are
  greatly simplified by leveraging the unification procedure provided by the meta language.
  Notable examples of meta languages are Elf~\cite{elf}, Twelf~\cite{twelf},
  $\lambda$Prolog~\cite{miller_nadathur_2012} and
  Isabelle~\cite{10.1007/978-3-540-71067-7_7}
  which have been %utilized
  used to implement %various
  many formal systems such as
  First Order Logic~\cite{felty88cade},
  Set Theory~\cite{10.1007/BF00881873},
  Higher Order Logic~\cite{books/sp/NipkowPW02}, and %even
  the Calculus of
  Constructions~\cite{felty93lics}.

  The object logic we are interested in is
  Coq's Calculus of Inductive Constructions (CIC)\cite{Coq-refman}. We aim to
  develop a higher-order unification-based 
  proof search procedure for it using the meta language
  Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
  Elpi's equational theory includes
  $\eta\beta$-equivalence and features a
  higher-order unification procedure \Ue restricted to the pattern
  fragment~\cite{miller92jsc}.
  Elpi offers an encoding of CIC suitable
  for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}
  but restricts \Ue to roughly first-order unification problems only.
  We refer to this basic encoding as \Fo.

  In this paper we propose a more well-behaved encoding called \Ho,
  and show how to translate unification problems from \Fo{}
  to corresponding ones in \Ho.
  Consequently, we derive \Uo{}, the higher-order unification procedure of \Fo{}
    that honours $\eta\beta$-equivalence (for CIC functions), addresses
  problems within the pattern fragment, and allows
  for the use of heuristics to deal with problems outside the
  pattern fragment. Moreover, as \Uo{} delegates most of the work to \Ue,
  it can be used to efficiently simulate a logic program in \Fo{} by
  taking advantage of
  unification-related optimizations of the meta language, such as clause indexing.

@@

\begin{proposition}[Compilation round trip]\label{prop:comprt}
  % \elpiIn{compile F T [] M [] L [] S} then 
  % if \elpiIn{T} $\in$ S
  If $\C{s}{t}{m}{l}$ and $l \in \linkStore$ and $m \in \mapStore$
  and
  $\sigma = \{ A \mapsto t\}$ and $\mapping{X}{A}{n} \in \mapStore$
  then $\D{\sigma}{\mapStore}{\linkStore}{\rho}$ and
  $\rho X \Eo \rho s$.
\end{proposition}

\noindent
The statement above roughly states that the translation from
\Fo{} to \Ho is not lossy, it can be reversed. The statement
is made convoluted by the fact that we hid the procedure to
decompile a term in favor of the one to decompile substitutions.
We omit to sketch the proof of this property for brevity.
